---
title: "Data_Analysis_Project"
author: "Isabelle Jensen, Sonia Vargas"
date: "11/16/24"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This project explores cast vote records (CVRs) from the Maryland 2024 primary elections to uncover trends, patterns, and insights into voter behavior. By examining these detailed voting records, we aim to answer several newsworthy questions about the electoral process, such as straight-ticket voting rates, undervoting in down-ballot races, and the influence of ballot styles on voter choices.

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
library(dplyr)
library(readxl)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary 
for data you are working with, and discuss any caveats or issues you discovered 
working with this data. 

This data is pretty huge, but this is a collection of all of the data if we were to want to load all of them.

```{r}
Alleghany <- read_excel("data/01_Allegany PP24 CVRs/01PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Alleghany")
AnneArundel <- read_excel("data/02_Anne Arundel PP24 CVRs/02PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Anne Arundel")
BaltimoreCity <- read_excel("data/03_Baltimore City PP24 CVRs/03PP24_Prov_CVR_SBE.xlsx") |>   mutate(County = "Baltimore City")
BaltimoreCounty <- read_excel("data/04_Baltimore County PP24 CVRs/04PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Baltimore County")
Calvert <- read_excel("data/05_Calvert PP24 CVRs/05PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Calvert")
Caroline <- read_excel("data/06_Caroline PP24 CVRs/06PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Caroline")
Carroll <- read_excel("data/07_Carroll PP24 CVRs/07PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Carroll")
Cecil <- read_excel("data/08_Cecil PP24 CVRs/08PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Cecil")
Charles <- read_excel("data/09_Charles PP24 CVRs/09PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Charles")
Dorchester <- read_excel("data/10_Dorchester PP24 CVRs/10PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Dorchester")
Frederick <- read_excel("data/11_Frederick PP24 CVRs/11PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Frederick")
Garrett <- read_excel("data/12_Garrett PP24 CVRs/12PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Garrett")
Harford <- read_excel("data/13_Harford PP24 CVRs/13PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Harford")
Howard <- read_excel("data/14_Howard PP24 CVRs/14PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Howard")
Kent <- read_excel("data/15_Kent PP24 CVRs/15PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Kent")
Montgomery <- read_excel("data/16_Montgomery PP24 CVRs/16PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Montgomery")
PG <- read_excel("data/17_Prince George's PP24 CVRs/17PP24_Prov_CVR_SBE.xlsx") |>
  mutate(County = "Prince George's")
QueenAnnes <- read_excel("data/18_Queen Annes PP24 CVRs/18PP24_Prov_CVR_SBE.xlsx") |>
  mutate(County = "Queen Anne's")
StMarys <- read_excel("data/19_St Marys PP24 CVRs/19PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "St Marys")
Somerset <- read_excel("data/20_Somerset PP24 CVRs/20PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Somerset")
Talbot <- read_excel("data/21_Talbot PP24 CVRs/21PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Talbot")
Washington <- read_excel("data/22_Washington PP24 CVRs/22PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Washington")
Wicomico <- read_excel("data/23_Wicomico PP24 CVRs/23PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Wicomico")
Worcester <- read_excel("data/24_Worcester PP24 CVRs/24PP24_Prov_CVR_SBE.xlsx") |> 
  mutate(County = "Worcester")
  
```

## Questions to explore
1. What is the proportion of Straight-Ticket Voting?
2. What are the Undervoting Rates in Down-Ballot Races?
3. Are there certain Precincts that behave similarly due to some aspects?
4. How much Ballot Splitting can we see in the cast vote records?
5. What is the Impact of Ballot Style on Voting Patterns in Local Races?

## Deliverable 2 (week 12) - 10 percent: 
For a quick overlook of a few interesting counties in MD: 
* Alleghany county has 10,596 observations of 22 vars, 
* Baltimore City has 98,232 obsercations of 78 vars,
* Frederick County has 50,434 observations of 27 vars
* Howard County has 64,723 observations of 29 vars
* MoCo has 99,999 observations of 72 vars
* PG county has 99,999 observations of 76 vars
* Talbot county has 7,932 observations of 23 vars

It looks like the more rural/red(?) counties have less on their ballots and Cast Vote records. It seems like there were places with more seats on their ballots. 
One thing about this data is that there is so much of it for each county, especially
those that are more populated. To analyze our questions we would likely have to select
just a few counties for our analysis for sake of time and computing power. 
For Codes or definitions, everything seems to be straightforward, and races we are
unfamiliar with could be researched. We will probably work off of state-wide races like the Senate race and try to answer our questions based on what we have. 

* Since a lot of these school boards are non partisan for the most part, we are going
to observe the split ticketing of Republican candidates voting for or against 1776 candidates
as discussed in (this article)[https://cnsmaryland.org/2024/11/08/md-conservatives-make-gains-in-school-board-races/] and researched by us. 
```{r School Board Split Tickets}
#### Frederick #### 
table(Frederick$`Board of Education (50000167144)`)
# "Democrat" endorsed candidates: Bokee, Lowe, Monier, King Wilson Sr
# 1776 supported candidates: Colt Morningstar Black, Jaime Kiersten Brennan
# (source)[https://www.frederickdemocrats.org/md-voting/upcoming-elections/board-education]
Frederick <- Frederick |> 
  mutate(REP_BOE = case_when(
    `Board of Education (50000167144)` == "Colt Morningstar Black (10022022.0000302)" ~ "REP",
    `Board of Education (50000167144)` == "Jaime Kiersten Brennan (10022022.0000119)" ~ "REP",
    TRUE ~ "Other"
  ),
  not_split = ifelse(str_detect(`Ballot Style`, "Rep") & REP_BOE == "REP" | str_detect(`Ballot Style`, "Dem") & REP_BOE != "REP" , 1 , 0)) # if they are REP and voted for one of the 1776 ed candidates OR Dem and did NOT vote for one of the 1776 candidates, they did NOT split their PRIMARY ballot
table(Frederick$not_split) 
# 34949 not split
# 15485 split from party candidates

Frederick_Rep <- Frederick |> 
  filter(str_detect(`Ballot Style`, "Rep"))
table(Frederick_Rep$not_split)
# 9127 Republicans split
# 11330 republicans did not split (voted for 1776 candidates)
11330/(11330+9127)*100 # 55% did not split

#### Washington #### 
table(Washington$`Board of Education (50000167344)`)
# 1776 candidates: Charles A. Burkett, Ashley McCusker, Victoria Beachley, April Ann Zentmeyer*
# "Dem" endorsed Candidates: Pieter Bickford, Stan Stouffer, Robin Merchant Spickler, Melissa Williams
# (source)[https://bluevoterguide.org/district/MD/Washington_County/21474/169]
Washington <- Washington |> 
  mutate(REP_BOE = case_when(
    `Board of Education (50000167344)` == "Charles A. Burkett (10031031.0000509)" ~ "REP",
    `Board of Education (50000167344)` == "Ashley McCusker (10031031.0000542)" ~ "REP",
    `Board of Education (50000167344)` == "Victoria Beachley (10031031.0000172)" ~ "REP",
    `Board of Education (50000167344)` == "Melissa Williams (10031031.0000656)" ~ "REP",
    TRUE ~ "Other"
  ),
  not_split = ifelse(str_detect(`Ballot Style`, "Rep") & REP_BOE == "REP" | 
                       str_detect(`Ballot Style`, "Dem") & REP_BOE != "REP" , 1 , 0)) # if they are REP and voted for one of the 1776 ed candidates OR Dem and did NOT vote for one of the 1776 candidates, they did NOT split their PRIMARY ballot
table(Washington$not_split) 
# 12965 not split
# 11545 split from party candidates

Washington_Rep <- Washington |> 
  filter(str_detect(`Ballot Style`, "Rep"))
table(Washington_Rep$not_split)
# 6737 Republicans did not split (voted for 1776 candidates) 
6737/(6737+7302)*100 # 48% Rep did not split
# 7302 Republicans did split


# conservative candidates: Joseph Marchio, Paul Harrison, Melissa Goshorn
#### Talbot #### 
table(Talbot$`Board of Education (2) (50000173444)`)
# 1776 candidates: Ann C. O'Connor, Kerry Gunshenan, Karla Wieland-Cherry

Talbot <- Talbot |> 
  mutate(REP_BOE = case_when(
    `Board of Education (2) (50000173444)` == "Ann C. O'Connor (10122127.0000735)" ~ "REP",
    TRUE ~ "Other"
  ),
  not_split = ifelse(str_detect(`Ballot Style`, "Rep") & REP_BOE == "REP" | 
                       str_detect(`Ballot Style`, "Dem") & REP_BOE != "REP" , 1 , 0)) # if they are REP and voted for one of the 1776 ed candidates OR Dem and did NOT vote for one of the 1776 candidates, they did NOT split their PRIMARY ballot
table(Talbot$not_split) 
# 4077 not split
# 3855 split from party candidates

Talbot_Rep <- Talbot |> 
  filter(str_detect(`Ballot Style`, "Rep"))
table(Talbot_Rep$not_split)
# 239 Republicans did not split (voted for 1776 candidates) 
239/(239+3701)*100 # 6% Rep did not split
# 3701 Republicans did split

#### Anne Arundel #### 

# Conservative, not 1776 candidates: Chuck Yocum, Latoya Nkongolo, Dawn Pullium
# (source)[https://www.somdnews.com/news/state/conservative-pac-dems-face-off-over-maryland-school-board-races/article_084d9d36-9537-11ef-9d33-03374e9ddf74.html] 
table(AnneArundel$`Board of Education (1) (50000176644)`)
table(AnneArundel$`Board of Education (3) (50000173844)`) # Yocum
table(AnneArundel$`Board of Education (4) (50000176744)`)
table(AnneArundel$`Board of Education (5) (50000176844)`) # Latoya

AnneArundel <- AnneArundel |> 
  mutate(REP_BOE = case_when(
    `Board of Education (3) (50000173844)` == "Chuck Yocum (10024024.0000151)" ~ "REP",
    `Board of Education (5) (50000176844)` == "LaToya Nkongolo (10008008.0000078)" ~ "REP",
    TRUE ~ "Other"
  ),
  not_split = ifelse(str_detect(`Ballot Style`, "Rep") & REP_BOE == "REP" | 
                       str_detect(`Ballot Style`, "Dem") & REP_BOE != "REP" , 1 , 0)) # if they are REP and voted for one of the 1776 ed candidates OR Dem and did NOT vote for one of the 1776 candidates, they did NOT split their PRIMARY ballot
table(AnneArundel$not_split) 
# 55953 not split
# 37613 split from party candidates

AnneArundel_Rep <- AnneArundel |> 
  filter(str_detect(`Ballot Style`, "Rep"))
table(AnneArundel_Rep$not_split)
# 3903 Republicans did not split (voted for 1776 candidates) 
3903/(3903+31555)*100 # 11% Rep did not split
# 31555 Republicans did split
```

ndervote Coding

Montgomery

```{r}
Montgomery |>
  filter(
    `DEM President of the United States (4) (500001630104)` == "undervote" |
    `DEM President of the United States (6) (500001630106)` == "undervote" |
    `DEM President of the United States (8) (500001630108)` == "undervote"
  ) |>
  summarize(
    `Race 4` = sum(`DEM President of the United States (4) (500001630104)` == "undervote", na.rm = TRUE),
    `Race 6` = sum(`DEM President of the United States (6) (500001630106)` == "undervote", na.rm = TRUE),
    `Race 8` = sum(`DEM President of the United States (8) (500001630108)` == "undervote", na.rm = TRUE)
  )
5 +	61+	269	
#335 undervotes for president

Montgomery |>
  summarize(
    `Total DEM (4)` = sum(!is.na(`DEM President of the United States (4) (500001630104)`)),
    `Total DEM (6)` = sum(!is.na(`DEM President of the United States (6) (500001630106)`)),
    `Total DEM (8)` = sum(!is.na(`DEM President of the United States (8) (500001630108)`))
  )
2329 + 12548 + 62687
#77564

335/77564
#0.4% of all democratic presidential votes were undervotes


Montgomery |>
  filter(
    `REP President of the United States (4) (500001630204)` == "undervote" |
    `REP President of the United States (6) (500001630206)` == "undervote" |
    `REP President of the United States (8) (500001630208)` == "undervote"
  ) |>
  summarize(
    `REP Race 4` = sum(`REP President of the United States (4) (500001630204)` == "undervote", na.rm = TRUE),
    `REP Race 6` = sum(`REP President of the United States (6) (500001630206)` == "undervote", na.rm = TRUE),
    `REP Race 8` = sum(`REP President of the United States (8) (500001630208)` == "undervote", na.rm = TRUE),
    `Total REP (4)` = sum(!is.na(`REP President of the United States (4) (500001630204)`)),
    `Total REP (6)` = sum(!is.na(`REP President of the United States (6) (500001630206)`)),
    `Total REP (8)` = sum(!is.na(`REP President of the United States (8) (500001630208)`)))
# JUST undervotes:
13	+ 0 +	348
# 361 undervotes for Rep pres
# TOTAL REP votes:
435 + 0 + 11988
# % undervote REP:
361/12423
# 2.9% undervotes for Rep president out of total votes

Montgomery |>
  filter(
    `DEM U.S. Senator (5000017641)` == "undervote" )|>
  summarize(
    Total = sum(`DEM U.S. Senator (5000017641)`  == "undervote", na.rm = TRUE),
  )
# 897 senate dem undervotes

Montgomery |>
  summarize(
    totalvotes=sum(!is.na(`DEM U.S. Senator (5000017641)` )))
# Total senate dem votes 77564	

897/77564	
# 1.2% undervote


Montgomery |>
  filter(
    `REP U.S. Senator (5000017642)` == "undervote" )|>
  summarize(
    Total = sum(`REP U.S. Senator (5000017642)`  == "undervote", na.rm = TRUE),
  )
#306 senate rep undervotes

Montgomery |>
  summarize(
    totalvotes=sum(!is.na(`REP U.S. Senator (5000017642)` )))
#Total senate rep votes 16976		

306/16976
#1.8% of all senate rep votes were undervotes
```

Undervote Howard

```{r}
Howard |>
  filter(
    `DEM President of the United States (3) (500001630103)`  == "undervote" ) |>
  summarize(
    Count = sum(`DEM President of the United States (3) (500001630103)`  == "undervote", na.rm = TRUE)
  )
# 231 undervotes for president

Howard |>
  summarize(
    Count = sum(!is.na( `DEM President of the United States (3) (500001630103)`))
  )
# 44726	total dem pres votes

231/44726
# 0.5% of all democratic presidential votes were undervotes


Howard |>
  filter(
    `REP President of the United States (3) (500001630203)`   == "undervote" ) |>
  summarize(
    Count = sum(`REP President of the United States (3) (500001630203)`   == "undervote", na.rm = TRUE)
  )
# 512 undervotes for president

Howard |>
  summarize(
    Count = sum(!is.na(  `REP President of the United States (3) (500001630203)` ))
  )
# 15901

512/15901
# 3.2% undervote for president rep candidates

Howard |>
  filter(
   `DEM U.S. Senator (5000017641)`  == "undervote" ) |>
  summarize(
    Count = sum(`DEM U.S. Senator (5000017641)`  == "undervote", na.rm = TRUE)
  )
# 967 undervotes for senate

Howard |>
  summarize(
    Count = sum(!is.na( `DEM U.S. Senator (5000017641)`))
  )
# 44726	

927/44726
# 2.1% undervote democrate senate race


Howard |>
  filter(
   `REP U.S. Senator (5000017642)`  == "undervote" ) |>
  summarize(
    Count = sum(`REP U.S. Senator (5000017642)`  == "undervote", na.rm = TRUE)
  )
# 261 undervotes for senate

Howard |>
  summarize(
    Count = sum(!is.na( `REP U.S. Senator (5000017642)`))
  )
# 15901		

261/15901
# 1.6% undervotes for rep senate
```


Frederick Undervotes
```{r}
Frederick |>
  filter(
    `DEM President of the United States (6) (500001630106)`   == "undervote" ) |>
  summarize(
    Count = sum( `DEM President of the United States (6) (500001630106)`   == "undervote", na.rm = TRUE)
  )
# 168 undervotes for president

Frederick |>
  summarize(
    Count = sum(!is.na(  `DEM President of the United States (6) (500001630106)` ))
  )
# 25772	total dem pres votes

168/25772
# 0.6% of all democratic presidential votes were undervotes


Frederick |>
  filter(
   `REP President of the United States (6) (500001630206)`    == "undervote" ) |>
  summarize(
    Count = sum( `REP President of the United States (6) (500001630206)`    == "undervote", na.rm = TRUE)
  )
# 491 undervotes for president

Frederick |>
  summarize(
    Count = sum(!is.na(  `REP President of the United States (6) (500001630206)`  ))
  )
# 20457		total rep pres votes

491/20457
# 2.4% of all republican presidential votes were undervotes


Frederick |>
  filter(
   `DEM U.S. Senator (5000017641)`   == "undervote" ) |>
  summarize(
    Count = sum( `DEM U.S. Senator (5000017641)`    == "undervote", na.rm = TRUE)
  )
# 471 undervotes for senates

Frederick |>
  summarize(
    Count = sum(!is.na(  `DEM U.S. Senator (5000017641)`  ))
  )
# 25772	total dem senate votes

471/25772
# 1.8% of all democratic senate votes were undervotes

Frederick |>
  filter(
   `REP U.S. Senator (5000017642)`      == "undervote" ) |>
  summarize(
    Count = sum( `REP U.S. Senator (5000017642)`      == "undervote", na.rm = TRUE)
  )
# 442	undervotes for senate

Frederick |>
  summarize(
    Count = sum(!is.na( `REP U.S. Senator (5000017642)`    ))
  )
# 20457	total rep senate votes

442/20457
# 2.2% of all republican senate votes were undervotes
```

Baltimore City Undervotes
```{r}
BaltimoreCity |>
  filter(
    `DEM President of the United States (2) (500001630102)` == "undervote" |
    `DEM President of the United States (7) (500001630107)`    == "undervote" ) |>
  summarize(
    `Race 2` = sum( `DEM President of the United States (2) (500001630102)` == "undervote", na.rm = TRUE),
    `Race 7` = sum( `DEM President of the United States (7) (500001630107)`  == "undervote", na.rm = TRUE)
  )
57+	1146	
# 1203 undervotes for dem president

BaltimoreCity |>
  summarize(
    `Total DEM (2)` = sum(!is.na(`DEM President of the United States (2) (500001630102)`)),
    `Total DEM (7)` = sum(!is.na(`DEM President of the United States (7) (500001630107)` ))
  )
8430+	85254
# 93684

1203/93684
# 1.3% of all democratic presidential votes were undervotes

BaltimoreCity |>
  filter(
   `REP President of the United States (2) (500001630202)` == "undervote" |
    `REP President of the United States (7) (500001630207)`    == "undervote" ) |>
  summarize(
    `Race 2` = sum( `REP President of the United States (2) (500001630202)` == "undervote", na.rm = TRUE),
    `Race 7` = sum( `REP President of the United States (7) (500001630207)`  == "undervote", na.rm = TRUE)
  )
0	+175		
# 175 undervotes for rep president

BaltimoreCity |>
  summarize(
    `Total DEM (2)` = sum(!is.na(`REP President of the United States (2) (500001630202)`)),
    `Total DEM (7)` = sum(!is.na(`REP President of the United States (7) (500001630207)` ))
  )
0	+ 4090

# 4090
175/4090
# 4.3% of all republican presidential votes were undervotes

BaltimoreCity |>
  filter(
  `DEM U.S. Senator (5000017641)`   == "undervote" ) |>
  summarize(
    Count = sum( `DEM U.S. Senator (5000017641)`    == "undervote", na.rm = TRUE)
  )
# 3088 undervotes for senates

BaltimoreCity |>
  summarize(
    Count = sum(!is.na( `DEM U.S. Senator (5000017641)`  ))
  )
# 93684	total dem senate votes

3088/93684
# 3.3% of all democratic senate votes were undervotes


BaltimoreCity |>
  filter(
  `REP U.S. Senator (5000017642)`  == "undervote" ) |>
  summarize(
    Count = sum( `REP U.S. Senator (5000017642)`   == "undervote", na.rm = TRUE)
  )
# 157 undervotes for rep senate

BaltimoreCity |>
  summarize(
    Count = sum(!is.na( `REP U.S. Senator (5000017642)` ))
  )
# 4548		total rep senate votes

157/4548
# 3.5% of all republican senate votes were undervotes
```




Undervotes Baltimore County
```{r}
BaltimoreCounty |>
  filter(
    `DEM President of the United States (1) (500001630101)` == "undervote" |
   `DEM President of the United States (2) (500001630102)`  == "undervote" |
   `DEM President of the United States (7) (500001630107)`   == "undervote"
  ) |>
  summarize(
    `Race 1` = sum( `DEM President of the United States (1) (500001630101)` == "undervote", na.rm = TRUE),
    `Race 2` = sum( `DEM President of the United States (2) (500001630102)` == "undervote", na.rm = TRUE),
    `Race 7` = sum(`DEM President of the United States (7) (500001630107)` == "undervote", na.rm = TRUE)
  )
18 +	243 +	70	
# 331 undervotes for president

BaltimoreCounty |>
  summarize(
    `Total DEM (1)` = sum(!is.na(`DEM President of the United States (1) (500001630101)`)),
    `Total DEM (2)` = sum(!is.na(`DEM President of the United States (2) (500001630102)`)),
    `Total DEM (7)` = sum(!is.na(`DEM President of the United States (7) (500001630107)`))
  )
3565	+ 48222	+ 15297
# 67084

331/67084
# 0.5 percent of all dem votes for president were undervotes

BaltimoreCounty |>
  filter(
   `REP President of the United States (1) (500001630201)`  == "undervote" |
  `REP President of the United States (2) (500001630202)`    == "undervote" |
   `REP President of the United States (7) (500001630207)`      == "undervote"
  ) |>
  summarize(
    `Race 1` = sum( `REP President of the United States (1) (500001630201)`  == "undervote", na.rm = TRUE),
    `Race 2` = sum( `REP President of the United States (2) (500001630202)`   == "undervote", na.rm = TRUE),
    `Race 7` = sum(`REP President of the United States (7) (500001630207)`    == "undervote", na.rm = TRUE)
  )
0+	471	+72	
# 543 undervotes for president

BaltimoreCounty |>
  summarize(
    `Total rep (1)` = sum(!is.na(`REP President of the United States (1) (500001630201)` )),
    `Total rep (2)` = sum(!is.na(`REP President of the United States (2) (500001630202)`  )),
    `Total rep (7)` = sum(!is.na(`REP President of the United States (7) (500001630207)` ))
  )

0	+ 21924 + 5837
# 27761

543/27761
# 2% of all votes for rep president was an undervote

BaltimoreCounty |>
  filter(
  `DEM U.S. Senator (5000017641)`    == "undervote" ) |>
  summarize(
    Count = sum( `DEM U.S. Senator (5000017641)`    == "undervote", na.rm = TRUE)
  )
# 1107 undervotes for senates

BaltimoreCounty |>
  summarize(
    Count = sum(!is.na( `DEM U.S. Senator (5000017641)`   ))
  )
# 67084	total dem senate votes

1107/67084
# 1.7% of all democratic senate votes were undervotes

BaltimoreCounty |>
  filter(
  `REP U.S. Senator (5000017642)`  == "undervote" ) |>
  summarize(
    Count = sum( `REP U.S. Senator (5000017642)`  == "undervote", na.rm = TRUE)
  )
# 530 undervotes for rep senate

BaltimoreCounty |>
  summarize(
    Count = sum(!is.na( `REP U.S. Senator (5000017642)`))
  )
# 32915	total rep senate votes

530/32915
# 1.6% of all republican senate votes were undervotes
```


Results of undervotes for each race and county
Montgomery County
President
dem 0.4%
rep 2.9%
senate
dem 1.2%
rep 1.8%


Howard County
President
dem 0.5%
rep 3.2%
senate
dem 2.1%
rep 1.6%

Frederick County
President
dem 0.6%
rep 2.4%
senate
dem 1.8%
rep 2.2%

Baltimore City
President
dem 1.3%
rep 4.3%
senate
dem 3.3%
rep 3.5%


Baltimore County
President
dem 0.5%
rep 2.0%
senate
dem 1.7%
rep 1.6%

Across the 5 counties Republicans had a higher undervote percentage with the average being 2.3% higher than Democrats. Republicans undervoted more than Democrats for Senate but the difference is much smaller (an average of 0.12%) This might show that Republicans in Maryland are less happy with their options for president in the primary than Democrats were. 

