---
title: "Class Reference"
author: "Sonia Vargas"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
# setwd("~/jour472/data_journalism_2024_fall")
getwd()
setwd("~/Documents/UMD/2024-2025/JOUR772_data_journalism")
```

### How to install and load the tidyverse

```{r}
# install.packages("tidyverse")
library(tidyverse)
```
```{r}
################# Three Functions Notes Ch 10 ##########################
?read_rds
# a tidyverse function that helps to read in data - im unsure ive ever used this format before, what does it stand for R dataset? it seems like its saving binary files from an origin (website) that can be used as data in the terminal
maryland_winred_contributions <- read_rds("https://thescoop.org/files/maryland_winred.rds")

?summarise
# summarise() creates a new data frame. It returns one row for each combination of grouping variables; if there are no grouping variables, the output will have a single row summarising all observations in the input. It will contain one column for each grouping variable and one column for each of the summary statistics that you have specified.
# this looks like its typically used when there is a grouped variable

?group_by
# Most data operations are done on groups defined by variables. group_by() takes an existing tbl and converts it into a grouped tbl where operations are performed "by group". ungroup() removes grouping.
# I personally really enjoy this function, it makes conceptualizing, rearranging, and analyzing data fun
maryland_winred_contributions %>% # i prefer piping but i never knew about the |> ability!
  group_by(memo_text) %>%  # this shows me the data by campaign contributions
  summarise(               # this is another var I want shown in this table
    count_contribs = n()   # this is the other variable, which counts n contributions to the group_by campaign titles
  )  # also with piping- a shortcut is cmd + shift + m

# wow! Hogan with top 4 campaign contributions


?arrange
# arrange() orders the rows of a data frame by the values of selected columns. 
# here im also going to take use of the basic summary statistic functions, mean(), min(), max(), sum()
maryland_winred_contributions %>%
  group_by(memo_text) %>%  # grouped by campaign contributions
  summarise(               # adding vars to table
    count_contribs = n(),   # counts n contributions to the group_by campaign titles
    total_amount = sum(amount), # adds up values of all contributions grouped by campaign
    mean_amount = mean(amount),
    min_amount = min(amount),
    max_amount = max(amount),
  ) %>% 
  arrange(desc(total_amount)) # I like changing what the arranging order variable is to see whats changing
# while Hogan is #4 for numbers of donations, he is #2 right behind trump for total contributions.

```

```{r}
################# Pre Lab 03 ##########################
?mutate()
# mutate() creates new columns that are functions of existing variables. It can also modify (if the name is the same as an existing column) and delete columns (by setting their value to NULL).

?clean_names()

# Resulting names are unique and consist only of the _ character, numbers, and letters. Capitalization preferences can be specified using the case parameter.

?rename()
# rename() changes the names of individual variables using new_name = old_name syntax; rename_with() renames columns using a function.

?get_dupes()
# For hunting duplicate records during data cleaning. Specify the data.frame and the variable combination to search for duplicates and get back the duplicated rows.

?distinct
# Keep only unique/distinct rows from a data frame. This is similar to unique.data.frame() but considerably faster.

?str_sub
# str_sub() extracts or replaces the elements at a single position in each string. str_sub_all() allows you to extract strings at multiple elements in every string.

?str_tot_title()
```

```{r}
############## other notes ###################
?melt
# base r
# The melt() function in R programming is an in-built function. It enables us to reshape and elongate the data frames in a user-defined manner. It organizes the data values in a long data frame format. kinda like a pivot longer!
# Syntax:
# melt(data-frame, na.rm = FALSE, value.name = “name”, id = 'columns')
```



