---
title: "Reverse Engineering Project"
author: "Luke Denbo, Isabelle Jensen, Sonia Vargas"
date: "10/19/24"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction 

In this notebook, we are reverse engineering the story, 
[Millions in out-of-state donations help fuel high-profile Maryland Democratic governor candidates]
(https://cnsmaryland.org/2022/03/03/millions-in-out-of-state-donations-help-fuel-high-profile-maryland-democratic-governor-candidates/)

This is the version of our project notebook with the data cleaning section implementing
case_when coding. In an earlier version we implemented a loop to detect states in the data. 
That version can be seen here [Reverse_Engineering_Older](https://github.com/s-vargas/data_journalism_2024_fall/blob/main/major_assignments/reverse_engineering/reverse_engineering_older.Rmd)
In this older version we created a dataframe of state abbreviations. Then, through
the loop a state column was created and would place the two-letter state abbreviation
as the column observation if one of the dataframe's abbrv was detected in the address string. 
It was efficient, but not exhaustive. In the newer version there was more work 
involved to find the inconsistent formatting in the addresses and correctly code those
states. However, this proved to be more accurate in finding which states had been 
coded incorrectly, and the amount of international donations that needed to be 
coded separately as well. Both methods were able to confirm the accuracy of the
engineered sentences from the article. 
## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
library(dplyr)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary 
for data you are working with, and discuss any caveats or issues you discovered 
working with this data. 

Data is provided by Professor Derek Willis. Each dataset is composed of the 
donations information for each candidate. 
This data has one column with an entire address of each donor. 
To successfully reverse engineer we need to tease out the states into its own col.
```{r}

# Load required data
Moore <- read_csv("data/Moore_ContributionsList.csv")
Perez <- read_csv("data/Perez_ContributionsList.csv")
King <- read_csv("data/King_ContributionsList.csv")
Franchot <- read_csv("data/Franchot_ContributionsList.csv")

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

```
```{r}

# Clean required data and prepare for analysis if needed. 

# This is so that we are able to organize donations by states, as detailed in the story.
# To do this, I am going to employ the following function.
?str_sub_all()
?str_detect
# all states are right before the zip code (-6 ch), and only have two letters (start at -8)

# We will check what states did not format correctly because of inconsistently 
# formatted zipcodes (ie. 20740-, 20740-0000 ), or from addresses without states at
# all so they are not where we expect the two letter combo to be.
```
```{r}
#### MOORE DATA CLEANING ####

# create the state column by isolating the 2 letter state abbreviation with str_sub
Moore  <-  Moore |> 
  mutate(state = str_sub(`Contributor Address`, start = -8, end = -7))

# lets table that and see what weird addresses didnt work with the str_sub
table(Moore$state)
# list of rogue addresses and finding out where the addresses should be indicated as
# 0 - 1 NORWAY
# 21 - 1 MD
# AE - 1 APO
# Al - 1 SP
# ap - 1 SINGAPORE
# be - 1 AUS
# D - 5 MD
# Du - 1 Dubai UAE
# g - 1  Hong Kong
# is- 1 Lisbon Portugal
# on - 1 UK
# tc - 1 UK

Moore |>  # this is to see the address of all of these outliers and find their true 'state'
  filter(
    state == " O" |
    state == "21" |
    state == "AE" |
    state == "Al" |
    state == "ap" |
    state == "be" |
    state == "D " |
    state == "Du" |
    state == "g " |
    state == "is" |
    state == "on" |
    state == "tc"
  ) |> 
  select(`Contributor Address`, state)

# cleaning up the rogue states and classifying them correctly
Moore <- Moore |> 
  mutate(state = case_when(
    state == " O" ~ "NOR",
    state == "D " ~ "MD",
    state == "21" ~ "MD",
    state == "SP" ~ "ES",
    state == "ap" ~ "SG",
    state == "be" ~ "AUS",
    state == "Du" ~ "UAE",
    state == "g " ~ "HK",
    state == "is" ~ "POR",
    state == "on" ~ "UK",
    state == "tc" ~ "UK",
    TRUE ~ state
  ))

# all clean!
table(Moore$state)
```
```{r}
#### PEREZ DATA CLEANING ####

# create the state column by isolating the 2 letter state abbreviation with str_sub
Perez  <-  Perez |> 
 mutate(state = str_sub(`Contributor Address`, start = -8, end = -7))

# lets table that and see what weird addresses didnt work with the str_sub
table(Perez$state)
# list of rogue addresses and finding out where the addresses should be indicated as
# 21 - 1
# AE - 1
# D - 7

# this is to see the address of all of these outliers and find their true 'state'
Perez |> 
  filter(state == "21"|
         state == "D " |
        state == "AE") |>
  select(`Contributor Address`, state)

# cleaning up the rogue states and classifying them correctly
Perez <- Perez |> 
  mutate(state = case_when(
    state == "21" ~ "MD",
    state == "D " ~ "MD",
    state == "AE" ~ "AP",
    TRUE ~ state
  ))

# all clean!
table(Perez$state)

```
```{r}
#### KING DATA CLEANING ####

# create the state column by isolating the 2 letter state abbreviation with str_sub
King  <-  King |>  
 mutate(state = str_sub(`Contributor Address`, start = -8, end = -7)) 


# lets table that and see what weird addresses didnt work with the str_sub
table(King$state)
# this is to see the address of all of these outliers and find their true 'state'
# D - 1 MD
# AP - 1 Military
# e - 2 Canada
# GU - 3 guam

# this is to see the address of all of these outliers and find their true 'state'
King |> 
  filter(state == "D "|
         state == "e ")

# cleaning up the rogue states and classifying them correctly
King <- King |> 
  mutate(state = case_when(
    state == "e " ~ "CAN",
    state == "D " ~ "MD",
    TRUE ~ state
  ))

# all clean!
table(King$state)

```
```{r}
#### FRANCHOT DATA CLEANING ####

# create the state column by isolating the 2 letter state abbreviation with str_sub
Franchot <-  Franchot |>  
 mutate(state = str_sub(`Contributor Address`, start = -8, end = -7)) 

# lets table that and see what weird addresses didnt work with the str_sub
table(Franchot$state)

# picking out the ones that did not come out right 
Franchot |> 
  filter(state == "00"|
         state == "01" |
         state == "03" |
         state == "04" |
         state == "05" |
         state == "06" |
         state == "07" |
         state == "08" |
         state == "09" |
         state == "11" |
         state == "13" |
         state == "14" |
         state == "15" |
         state == "16" |
         state == "18" |
         state == "20" |
         state == "21" |
         state == "22" |
         state == "23" |
         state == "24" |
         state == "33" |
         state == "40" |
         state == "60" |
         state == "65" |
         state == "70" |
         state == "71" |
         state == "74" |
         state == "75" |
         state == "77" |
         state == "78" |
         state == "79" |
         state == "80" |
         state == "81" |
         state == "85" |
         state == "87" |
         state == "92" |
         state == "93" |
         state == "95" |
         state == "A " |
         state == "D " |
         state == "C " |
         state == "E " |
         state == "L " |
        state == "Y ") |> 
  select(`Contributor Address`, state)

# isolating ones that have more than 1 observation to see if it is 1 or more distinct states
Franchot |>  
  filter(state == "00"|
         state == "01" |
         state == "18" |
         state == "21" |
         state == "33" |
         state == "70" |
         state == "93" |
         state == "95" |
         state == "A " |
         state == "D " |
         state == "C " |
         state == "E " |
         state == "L " |
        state == "Y ") |> 
  select(`Contributor Address`, state) |> 
  arrange(desc(state))
# The following that have multiple observations AND multiple unique states
Franch_odds <- Franchot |>  # the duplicates that have multiple states
  filter(state == "00"|
         state == "01" |
         state == "21" |
         state == "70" |
         state == "A ") |> 
  select(`Contributor Address`, state) |> 
  arrange(desc(state))

Franchot <- Franchot |> 
  mutate(state = str_sub(`Contributor Address`, start = -8, end = -7)) |> 
  # first doing the str_sub for the ones that have uniform address zip format
  mutate(state = case_when( 
    # cleaning the ones that didnt format right because of different address format
    state == "00" ~ "MD", # also DC
    state == "01" ~ "MD", # also VA
    state == "03" ~ "MD",
    state == "04" ~ "MD",
    state == "05" ~ "MD",
    state == "06" ~ "MD",
    state == "07" ~ "MD",
    state == "08" ~ "MD",
    state == "09" ~ "MD",
    state == "11" ~ "MD",
    state == "13" ~ "MD",
    state == "14" ~ "MD",
    state == "15" ~ "MD",
    state == "16" ~ "MD",
    state == "18" ~ "VA",
    state == "20" ~ "MD",
    state == "21" ~ "MD", # and FL
    state == "22" ~ "MD",
    state == "23" ~ "MD",
    state == "24" ~ "MD",
    state == "33" ~ "PA",
    state == "40" ~ "MD",
    state == "60" ~ "MD",
    state == "65" ~ "MD",
    state == "70" ~ "MD", # also NJ
    state == "71" ~ "MD",
    state == "74" ~ "MD",
    state == "75" ~ "MD",
    state == "77" ~ "MD",
    state == "78" ~ "MD",
    state == "79" ~ "MD",
    state == "80" ~ "MD",
    state == "81" ~ "MD",
    state == "85" ~ "MD",
    state == "87" ~ "MD",
    state == "92" ~ "MD",
    state == "93" ~ "WY",
    state == "95" ~ "FL",
    state == "A " ~ "VA", # also PA, MA, VA
    state == "D " ~ "MD",
    state == "C " ~ "DC",
    state == "E " ~ "DE",
    state == "L " ~ "FL",
    state == "Y " ~ "NY",
    TRUE ~ state)) |>
  mutate(state = case_when(
    # finally, assigning the correct states to the leftovers that had duplicates 
    # from the original state assignment method with str_sub
    `Contributor Address` == "1735 Market Street  51st Floor  Philadelphia  PA 19103-" ~ "PA",
    `Contributor Address` == "1950 Old Gallows Road #600  Vienna  VA 22182-" ~ "VA",
    `Contributor Address` == "30 Main Street  Yarmouth Port  MA 02675-" ~ "MA",
    `Contributor Address` == "95 Oxford Avenue  Cambridge  MA 06671-" ~ "MA",
    `Contributor Address` == "15 America Ave.  Suite 302  Lakewood  NJ 08701-4594" ~ "NJ",
    `Contributor Address` == "4306 Lake Woodbourne Dr.  Jacksonville  FL 32217-4482" ~ "FL",
    `Contributor Address` == "5818 Robins Nest Ln.  Burke  VA 22015-3118" ~ "VA",
    `Contributor Address` == "733 - 15th St.  N.W.  Apt. 719  Washington  DC 20005-2163" ~ "DC",
    `Contributor Address` == "500 Eighth Street  NW  Washington  DC 20004-2131" ~ "DC",
    TRUE ~ state
  ))

# checking to see if most of the addresses changed to the correct state 

Franchot |> 
  filter(state == "PA" | state == "VA" | state == "MA" | state == "FL" | state == "NJ") |> 
    select(`Contributor Address`, state) |> 
  arrange(state)

table(Franchot$state) # all clean!
```
## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: “Moore raised over $1 million in New York, and Perez gained 
over $226,000 from the state. Perez also received nearly $235,000 from California 
donors and raised over $100,000 in Florida and Virginia. Moore and King collected 
over $100,000 each from Illinois.”

* **Analysis summary**: 
* For the Moore dataset, it is confirmed that he raised over 
$1 million in New York, and over $100,000 in Illinois. 
* For the Perez data, it is confirmed that he raised $226,208 in New York, almost
$235,000 in California, and over $100,000 in each Florida and Virignia.
* For the King data, it is confirmed that He raised over $100,000 from Illinois. 
* All of the data in this paragraph has been confirmed. After singling out the state
through str_sub function, all data has checked out as correctly represented in the story. 

```{r}
# Put code to reverse engineer sentence here

# Moore data: New York, Illinois
Moore |> 
  filter(state == "NY" | state == "IL") |> 
  group_by(state) |> 
  summarise(
    total = sum(`Contribution Amount`)
  ) |> 
  arrange(desc(total))
# we see here that Moore raised exactly $1,002,915 from NY, and $131,629 from IL.

# Perez data: New York, California, Florida, Virginia
Perez |> 
  filter(state == "NY" | state == "CA" | state == "FL" | state == "VA") |> 
  group_by(state) |> 
  summarise(
    total = sum(`Contribution Amount`)
  ) |> 
  arrange(desc(total))
# we see here that Perez raised exactly $226,208.46 from NY, 
# $234,749 from California, $123,178 from VA, and $110,639 from FL.

# King data: Illinois
King |> 
  filter(state == "IL") |> 
  group_by(state) |> 
  summarise(
    total = sum(`Contribution Amount`)
  ) |> 
  arrange(desc(total))
# King raised $125,055.4 in Illinois

# Display results of code below this codeblock

```

### Sentence 2

* **Sentence text**: Moore, a U.S. Army veteran and bestselling author, built a 
campaign war chest that included over $2.3 million in out-of-state contributions. 
Tom Perez, a former leader of the Democratic National Committee, raised nearly 
$1.7 million in donations from outside Maryland.
* **Analysis summary**: I was able to confirm both findings. Moore was found to 
have raised over $\$2.3$ million from out of state, and Perez almost $\$1.7$ million 
as the story states.

```{r}
# Put code to reverse engineer sentence here
Moore |> 
  filter(state != "MD") |> 
  summarise(
    M_total = sum(`Contribution Amount`)
  )
# 2,387,166	

Perez |> 
  filter(state != "MD") |> 
  summarise(
    P_total = sum(`Contribution Amount`)
  )
# 1,695,893

# Display results of code below this codeblock

```

### Sentence 3

* **Sentence text**: John B. King, former secretary of Education under President 
Barack Obama, raised $2.2 million in campaign funding last year. Only 11%, just 
under $240,000, came from Maryland donors. King raised over $534,000 in New York alone.
* **Analysis summary**: King raised a total of $~\$2.2$ million. $\$238,527$ came 
from MD, and $\$534,478$ came from NY. The ratio of MD donations is $~11\%$ of total 
donations

```{r}
# Put code to reverse engineer sentence here

King |> 
    summarise(
    K_total = sum(`Contribution Amount`)
  )
# Raised a total of $2,199,015
King |> 
  filter( state == "MD" | state == "NY") |> 
  group_by(state) |> 
    summarise(
    K_total = sum(`Contribution Amount`)
  )
# MD donations: $238527; NY donations: $534,478
238527/2199015 
# Display results of code below this codeblock

```

### Sentence 4

* **Sentence text**: Moore, King and Perez stand in stark contrast to Peter Franchot, 
a former Maryland state delegate and current comptroller, who raised over 78% of 
his $1.66 million from inside Maryland.
* **Analysis summary**: This is confirmed, as Franchot raised $\$1,306,371$ of 
his total $\$1,663,516$ contributions from Maryland. This comes out to 78% of 
total contributions from Maryland. Unsure why they used the language for "over 
78%" when it comes out to 78.3% according to this data and how it is coded here. 

```{r}
# Put code to reverse engineer sentence here
Franchot |> 
  summarise(total = sum(`Contribution Amount`))
# $1,663,516

Franchot |> 
  filter( state == "MD") |> 
  summarise(total = sum(`Contribution Amount`))
# MD contributions: 1,306,371
1306371/1663516
# 78%

# Display results of code below this codeblock

```

### Sentence 5

* **Sentence text**: Less than a third of the money Perez raised, 31%, came from 
Maryland donors.
* **Analysis summary**: Perez raised a total of $\$2,458,688$, with $\$762,794.4$
coming from MD. This comes out to 31% of his donations coming from Maryland donors,
and aligns with the story. 

```{r}
# Put code to reverse engineer sentence here
Perez |>  
  summarise(
    total = sum(`Contribution Amount`)
  )
# 2,458,688 total donations
Perez |> 
  filter(state == "MD") |> 
  summarise(
    total = sum(`Contribution Amount`)
  )
# $762,794.4 from MD
762794/2458688
#31.0% come from MD

# Display results of code below this codeblock

```