---
title: "Reverse Engineering Project"
author: "Luke Denbo, Isabelle Jensen, Sonia Vargas"
date: "9/28/24"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, 
[Millions in out-of-state donations help fuel high-profile Maryland Democratic governor candidates]
(https://cnsmaryland.org/2022/03/03/millions-in-out-of-state-donations-help-fuel-high-profile-maryland-democratic-governor-candidates/)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary 
for data you are working with, and discuss any caveats or issues you discovered 
working with this data. 

Data is provided by Professor Derek Willis. Each dataset is composed of the 
donations information for each candidate. 
This data has one column with an entire address of each donor. 
To successfully reverse engineer we need to tease out the states into its own col.
```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(dplyr)

# Load required data
setwd("~/UMD/2024-2025/data_journalism_2024_fall/major_assignments/reverse_engineering/data")

Moore <- read_csv("data/Moore_ContributionsList.csv")
Perez <- read_csv("data/Perez_ContributionsList.csv")
King <- read_csv("data/King_ContributionsList.csv")
Franchot <- read_csv("data/Franchot_ContributionsList.csv")

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 

# This is so that we are able to organize donations by states, as detailed in the story.
# To do this, I am going to employ the following function.
?str_sub_all()
?str_detect

# all states are right before the zip code (-6 ch), and only have two letters (start at -8)
# Moore  <-  Moore |> 
#   mutate(state = str_sub(`Contributor Address`, start = -8, end = -7))
# 
# table(Moore$state)
# 

# Perez  <-  Perez |> 
#  mutate(state2 = str_sub(`Contributor Address`, start = -8, end = -7))
# 
# King  <-  King |>  
#  mutate(state = str_sub(`Contributor Address`, start = -8, end = -7)) 
# this address data is a lot messier than i thought, going to attempt a str_detect method

states <- data.frame(
  state_abbrv = c(" AK"," AL"," APO"," AR"," AZ"," CA"," CO"," CT"," DC"," DE"," FL"," GA",
                     " GU"," HI"," IA"," ID"," IL"," IN"," KS"," KY"," LA"," MA"," MD",
                     " ME"," MI"," MN"," MO"," MS"," MT"," NC"," ND"," NE"," NH"," NJ"," NM",
                     " NV"," NY"," OH"," OK"," OR"," PA"," PR"," RI"," SC"," SD"," TN"," TX",
                     " UT"," VA"," VI"," VT"," WA"," WI"," WV"," WY", "YT CA"))
# 7 NON-STATES: AP(Armed Forces Pacific), PR(Puerto Rico), GU(Guam), VI(Virgin Islands), DC
# CANADA: YT CA
# the spaces here are very important, otherwise it will detect combinations of the
# letters from within the address names

for (state in states$state_abbrv) {
  Moore$state[str_detect(Moore$`Contributor Address`, state)] <- state
}

for (state in states$state_abbrv) {
  Perez$state[str_detect(Perez$`Contributor Address`, state)] <- state
}

for (state in states$state_abbrv) {
  King$state[str_detect(King$`Contributor Address`, state)] <- state
}

for (state in states$state_abbrv) {
  Franchot$state[str_detect(Franchot$`Contributor Address`, state)] <- state
}

table(Moore$state)
table(Perez$state)
table(King$state)
table(Franchot$state)
```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: “Moore raised over $1 million in New York, and Perez gained 
over $226,000 from the state. Perez also received nearly $235,000 from California 
donors and raised over $100,000 in Florida and Virginia. Moore and King collected 
over $100,000 each from Illinois.”

* **Analysis summary**: 
* For the Moore dataset, it is confirmed that he raised over 
$1 million in New York, and over $100,000 in Illinois. 
* For the Perez data, it is confirmed that he raised $226,208 in New York, almost
$235,000 in California, and over $100,000 in each Florida and Virignia.
* For the King data, it is confirmed that He raised over $100,000 from Illinois. 
* All of the data in this paragraph has been confirmed. After singling out the state
through str_sub function, all data has checked out as correctly represented in the story. 

```{r}
# Put code to reverse engineer sentence here

# Moore data: New York, Illinois
Moore |> 
  filter(state == "NY" | state == "IL") |> 
  group_by(state) |> 
  summarise(
    total = sum(Contribution.Amount)
  ) |> 
  arrange(desc(total))
# we see here that Moore raised exactly $1,002,915 from NY, and $131,629 from IL.

# Perez data: New York, California, Florida, Virginia
Perez |> 
  filter(state == "NY" | state == "CA" | state == "FL" | state == "VA") |> 
  group_by(state) |> 
  summarise(
    total = sum(Contribution.Amount)
  ) |> 
  arrange(desc(total))
# we see here that Perez raised exactly $226,208.46 from NY, 
# $234,749 from California, $123,178 from VA, and $110,639 from FL.

# King data: Illinois
King |> 
  filter(state == "IL") |> 
  group_by(state) |> 
  summarise(
    total = sum(Contribution.Amount)
  ) |> 
  arrange(desc(total))
# King raised $125,055.4 in Illinois



# Display results of code below this codeblock

```

### Sentence 2

* **Sentence text**: Moore, a U.S. Army veteran and bestselling author, built a 
campaign war chest that included over $2.3 million in out-of-state contributions. 
Tom Perez, a former leader of the Democratic National Committee, raised nearly 
$1.7 million in donations from outside Maryland.
* **Analysis summary**: I was able to confirm both findings. Moore was found to 
have raised over $\$2.3$ million from out of state, and Perez almost $\$1.7$ million 
as the story states.

```{r}
# Put code to reverse engineer sentence here
Moore |> 
  filter(state != " MD") |> # space is important cause thats how I coded the states
  summarise(
    M_total = sum(`Contribution Amount`)
  )
# 2,417,225	

Perez |> 
  filter(state != " MD") |> 
  summarise(
    P_total = sum(`Contribution Amount`)
  )
# 1,697,893

# Display results of code below this codeblock

```

### Sentence 3

* **Sentence text**: John B. King, former secretary of Education under President 
Barack Obama, raised $2.2 million in campaign funding last year. Only 11%, just 
under $240,000, came from Maryland donors. King raised over $534,000 in New York alone.
* **Analysis summary**: King raised a total of $~\$2.2$ million. $\$238,339$ came 
from MD, and $\$534,478$ came from NY. The ratio of MD donations is $~11\%$ of total 
donations

```{r}
# Put code to reverse engineer sentence here

King |> 
    summarise(
    K_total = sum(`Contribution Amount`)
  )
# Raised a total of $2,199,015
King |> 
  filter( state == " MD" | state == " NY") |> 
  group_by(state) |> 
    summarise(
    K_total = sum(`Contribution Amount`)
  )
# MD donations: $238339; NY donations: $534,478
238339/2199015 
# Display results of code below this codeblock

```

### Sentence 4

* **Sentence text**: Moore, King and Perez stand in stark contrast to Peter Franchot, 
a former Maryland state delegate and current comptroller, who raised over 78% of 
his $1.66 million from inside Maryland.
* **Analysis summary**: This is confirmed, as Franchot raised $\$1,304,121$ of 
his total $\$1,663,516$ contributions from Maryland. This comes out to 78% of 
total contributions from Maryland. Unsure why they used the language for "over 
78%" when it comes out to 78.4% according to this data and how it is coded here. 

```{r}
# Put code to reverse engineer sentence here
Franchot |> 
  summarise(total = sum(`Contribution Amount`))
# $1,663,516

Franchot |> 
  filter( state == " MD") |> 
  summarise(total = sum(`Contribution Amount`))
# MD contributions: 1,304,121
1304121/1663516
# 78%

# Display results of code below this codeblock

```

### Sentence 5

* **Sentence text**: Less than a third of the money Perez raised, 31%, came from 
Maryland donors.
* **Analysis summary**: Perez raised a total of $\$2,458,688$, with $\$760,794.4$
coming from MD. This comes out to 31% of his donations coming from Maryland donors,
and aligns with the story. 

```{r}
# Put code to reverse engineer sentence here
Perez |>  
  summarise(
    total = sum(`Contribution Amount`)
  )
# 2,458,688 total donations
Perez |> 
  filter(state == " MD") |> 
  summarise(
    total = sum(`Contribution Amount`)
  )
# $760,794.4 from MD
760794/2458688
#30.9% come from MD

# Display results of code below this codeblock

```

-30-