---
title: "Reverse Engineering Project"
author: "Luke Denbo, Isabelle Jensen, Sonia Vargas"
date: "9/28/24"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, 
[Millions in out-of-state donations help fuel high-profile Maryland Democratic governor candidates]
(https://cnsmaryland.org/2022/03/03/millions-in-out-of-state-donations-help-fuel-high-profile-maryland-democratic-governor-candidates/)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary 
for data you are working with, and discuss any caveats or issues you discovered 
working with this data. 

Data is provided by Professor Derek Willis. Each dataset is composed of the 
donations information for each candidate. 
This data has one column with an entire address of each donor. 
To successfully reverse engineer we need to tease out the states into its own col.
```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(dplyr)

# Load required data
setwd("~/UMD/2024-2025/data_journalism_2024_fall/major_assignments/reverse_engineering/data")

Moore <- read_csv("data/Moore_ContributionsList.csv")
Perez <- read_csv("data/Perez_ContributionsList.csv")
King <- read_csv("data/King_ContributionsList.csv")

# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 

# This is so that we are able to organize donations by states, as detailed in the story.
# To do this, I am going to employ the following function.
?str_sub_all()
# all states are right before the zip code (-6 ch), and only have two letters (start at -8)
Moore  <-  Moore |> 
  mutate(state = str_sub(Contributor.Address, start=-8, end=-7))

Perez  <-  Perez |> 
  mutate(state = str_sub(Contributor.Address, start=-8, end=-7))

King  <-  King |> 
  mutate(state = str_sub(Contributor.Address, start=-8, end=-7))

```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: “Moore raised over $1 million in New York, and Perez gained 
over $226,000 from the state. Perez also received nearly $235,000 from California 
donors and raised over $100,000 in Florida and Virginia. Moore and King collected 
over $100,000 each from Illinois.”

* **Analysis summary**: 
* For the Moore dataset, it is confirmed that he raised over 
$1 million in New York, and over $100,000 in Illinois. 
* For the Perez data, it is confirmed that he raised $226,208 in New York, almost
$235,000 in California, and over $100,000 in each Florida and Virignia.
* For the King data, it is confirmed that He raised over $100,000 from Illinois. 
* All of the data in this paragraph has been confirmed. After singling out the state
through str_sub function, all data has checked out as correctly represented in the story. 

```{r}
# Put code to reverse engineer sentence here

# Moore data: New York, Illinois
Moore |> 
  filter(state == "NY" | state == "IL") |> 
  group_by(state) |> 
  summarise(
    total = sum(Contribution.Amount)
  ) |> 
  arrange(desc(total))
# we see here that Moore raised exactly $1,002,915 from NY, and $131,629 from IL.

# Perez data: New York, California, Florida, Virginia
Perez |> 
  filter(state == "NY" | state == "CA" | state == "FL" | state == "VA") |> 
  group_by(state) |> 
  summarise(
    total = sum(Contribution.Amount)
  ) |> 
  arrange(desc(total))
# we see here that Perez raised exactly $226,208.46 from NY, 
# $234,749 from California, $123,178 from VA, and $110,639 from FL.

# King data: Illinois
King |> 
  filter(state == "IL") |> 
  group_by(state) |> 
  summarise(
    total = sum(Contribution.Amount)
  ) |> 
  arrange(desc(total))
# King raised $125,055.4 in Illinois



# Display results of code below this codeblock

```

### Sentence 2

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 3

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 4

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

### Sentence 5

* **Sentence text**: [Paste in sentence to engineer here]
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here

# Display results of code below this codeblock

```

-30-