---
title: "lab_04"
author: "Derek Willis"
date: "2024-09-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(janitor)
library(lubridate)
library(dplyr)

```

For this lab, we want to investigate spending by the two leading Senate candidates in Maryland during this election, Angela Alsobrooks and Larry Hogan. Our goal is to try and standardize the purpose of the expenditures so that we can compare the two campaigns. We have the data, but it's messy - purpose descriptions are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of campaign expenditures located in the data folder called "md_senate_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table
setwd("~/UMD/2024-2025/data_journalism_2024_fall/labs/lab_04")
expenses <- read.csv("md_senate_expenses.csv")
```

## Answer questions

**Q1.** You've been assigned to compare the spending priorities of the two campaigns, but they describe their spending differently using the `purpose` column. Using OpenRefine, create a project using the original CSV file, make a copy of the `purpose` column called `purpose_clean` and then standardize it, focusing on the purposes that appear most often or have the largest total amounts. You ccan transform the data as you want - making all records the same case, for example, to help. The goal is to be able to group together as many similar purpose values, so you can choose to call all transportation-related expenses the same thing.

How do you know when you're done? It's a judgment call, but you should be close to 100 unique values for `purpose_clean`, and maybe even below that number.

Then, when you are done standardizing `purpose_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.

Then, using that `purpose_clean` column, write code to find the largest amounts of spending on categories by both campaigns.

**A1.**

```{r}
expense_cleaner <- read.csv("md-senate-expenses-cleaner.csv")

expense_cleaner |> 
  group_by(candidate, purpose_clean) |> 
  summarise(
            total = sum(amount)
            ) |> 
  arrange(desc(total))
```

**Q2.** Let's do some research on the top purposes from A1. Feel free to search the Web to understand those terms better if need be, and then write a description comparing and contrasting the two campaigns' largest categories of spending.

**A2.**
Media advertising for Alsobrooks, and ADS for Hogan - it seems that both campaigns'
largest expense is advertising. Salary for Alsobrooks and fundraising for Hogan- 
Alsobrooks spends more on paying employees as second largest expense, Hogan spends 
money on fundraising activities.  

**Q3.** Pick one of the top categories for each campaign and write code that shows how that spending has changed over time, if at all. You'll need to use lubridate for this. Write a couple of sentences describing that pattern.

**A3.**
* The advertising spending from the Alsobrooks campaign barely existed in Dec 2023.
February saw the beginning of some more spending with $255,090, more than doubling
in March with $610,954, and peaking in April with $2,196,550. This shows that the bulk
of the spending ocurred in the month before the MD primary.
* The ADS spending from the Hogan campaign follows a similar pattern, except Hogan
shows a peak in May, the month of the primary. Both candidates taper off, but 
still have high spending in the month after their peak spending. Both spending around half a million. 

```{r}
expense_cleaner$date <- ymd(expense_cleaner$date)

expense_cleaner |> 
  filter(purpose_clean == "Media Advertising") |> 
  mutate(month = floor_date(date, "month")) |>
  group_by(month) |> 
   summarise(
    total = sum(amount)
  ) |>
  arrange(month)

expense_cleaner |> 
  filter(purpose_clean == "ADS") |> 
  mutate(month = floor_date(date, "month")) |>
  group_by(month) |> 
   summarise(
    total = sum(amount)
  ) |>
  arrange(month)

```

**Q4.** Which committee has the larger payroll or salary spending so far? How many unique recipients of payroll/staff spending are there for each candidate?

**A4.**
* Alsobrooks has the larger payroll/salary with $822,015 in this dataset. Who gets paid,
all individuals, are listed in the dataset.
* Hogan seems to fund an organization called 'Purple Surfboard', headquartered in Annapolis.
After some quick googling it seems that this is a term coined by Hogan after a successful
reelection campaign- similar to "blue wave" terminology. I would believe that this is
his registered company that staff/employees get paid through. He does seem to
pay less than Alsobrooks, with $481,109. He does, however have a longer history with state politics
and may be able to be more efficient with staff spending. 
```{r}
# unique names
expense_cleaner$names <- paste(expense_cleaner$last_name, " ", expense_cleaner$first_name)

# Alsobrooks
expense_cleaner |> 
  filter(purpose_clean == "Salary") |> 
  group_by(names) |> 
   summarise(
    total = sum(amount)
  ) |> 
  arrange(desc(total))
# $822,015 total overall

# Hogan
expense_cleaner |> 
  filter(purpose_clean == "STAFFING - SEE MEMO" |purpose_clean ==  "STAFFING SERVICES")

expense_cleaner |> 
  filter(purpose_clean == "STAFFING - SEE MEMO" |purpose_clean ==  "STAFFING SERVICES") |> 
  group_by(organization_name) |> 
   summarise(
    total = sum(amount)
  ) |> 
  arrange(desc(total))
# $481,109 total overall
```

**Q5** What's the most interesting story idea you've found in this data?

**A5**
* I think the differences in staffing and salaries that are paid by each campaign is pretty interesting. 
Im sure this has something to do with Alsobrooks needing to build up a campaign team
to this scale for the first time. 
* I am also interesting in knowing what the exact purposes
of all of the car spending are for Hogan as funded by his campaign spending. With a 
statewide campaign it would make sense to cover costs related to car travel, but
more research into that specifically would be cool. 
* When it comes to travel, Alsobrooks spent $65,189.83, and Hogan spent $55,343.51.
These are similar numbers, and the data is for before and directly after the primary.
Still, I wonder how with the nationalization of a campaign like Hogan's, why they have similar
travel spending. Does Hogan take advantage of his national popularity? Does he travel outside of MD?